Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-05 Manuscript/manuscript.Rmd
wip on intro

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones associated with TAD boundaries still leaves much to be desired.

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, this is a multi-faceted probelem. Many methods ignore key characteristics [?? which ones ??; !!!editted] of 3D genomic data. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->

Firstly, sparsity of TAD boundary regions throughout the linear genome contribute to heavily imbalanced classes. This imbalance results in small imbalance ratios. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing] [?? This sentence is very difficult to understand. Don't see connection between the bias and the error rate ??; !!!].

We can account for this imbalance at the data level in addition to at the evaluation level [?? I rewrote this sentence. Was very wordy and difficult to follow ??]. Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. Evaluation level solutions include choosing suitable performance metrics for evaluating models. <!--These solutions are necessary given that the test set, used to validate re-sampled training sets, is itself imbalanced.--> It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has been studied previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<!--
In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. The class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. 
-->

<!--
A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.
-->

<!--
Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role]. 
-->

Secondly, it is important to determine the optimal representation of the relationship between features in the model and genomic regions. This is known as feature engineering. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e.g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

<!--
[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).
-->

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling [@chawla2002smote]. Under-sampling is performed without replacement from the majority class. Over-sampling is performed by creating new synthetic observations using the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the feature vectors joining any/all of the k=5 (default) minority class nearest neighbors. The SMOTE algorithm can have many variations in the weightinh of over- and under-sampling. In this study we restricted the algorithm to 100% over-sampling and 200% under-sampling. That is, 1 new synthetic observation was to be created for each existing minority class data point, while 2 data points were simultaneously randomly selected from the majority class. This combination results in perfectly balanced classes. The *DMwR* R package was used to perform SMOTE (version 0.4.1).

@@

## DATA CONSTRUCTION FOR UNBIASED COMPARISONS BETWEEN RESOLUTION-SPECIFIC RANDOM FOREST CLASSIFIERS
<!--## CELL LINE AND DATA [??? What data?] RESOLUTION SPECIFIC MODEL [??? What model?] CONSTRUCTION [??? Why?] !!!created new header-->

In total, we analyzed publically available Hi-C data for both the GM12878 and K562 cell lines, at four resolutions including 10 kb, 25 kb, 50 kb, and 100 kb. Given the varying levels of data resolution, we created a resolution-specific approach toward data constuction used for downstream modelling (Figure 1). This allowed us an unbiased way to compare performances of random forest classifiers between varying resolutions of Hi-C data. [??? Boring. What problem you are solving?;!!!editted]

@@

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD regions, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at each of the four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). However, imbalance ratios were found to be invariant to resolution, remaining consistently low with very little variability between chromosomes, regardless of cell line.

**Table 1. Sparce TAD boundaries throughout the linear genome create high imbalance ratios irrespective of resolution.**

@@

<!--We then went on the evaluate the effect of class imbalance on model performance.--> In order to set a baseline in performance, we first evaluated the least granular random forest classifier, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the complexity in data level characteristics by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).

>>>>

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [??? REFS? Carefully select from my paper notes]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. [??? This sentence and reference belong to the previous paragraph, think and move] It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 <!--have been shown to -->colocalize <!--together [??? Note removal of "water" words]-->near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d]. [??? This sentence and reference belong to the previous paragraph, think and move] However, the growing number of annotated genomic features and the methods used to identify the ones associated with TAD boundaries still leaves much to be desired. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it.]

[!!! This paragraph defines problems]

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remains unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced ratio of positive/negative classes usef to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, it remains unclear how the increased resolution of Hi-C data affect the performance of predictive modeling. Lastly, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored.

[!!! The following paragraph should review previous work and highlight pitfalls, creating motivation for yor study]

<!--Secondly, it is important to determine the optimal representation of the relationship between features in the model and genomic regions. This is known as feature engineering. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e.g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. Here, using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. Moreover, while direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving forces of genome regulation [??? REFS]. These relationships can be quantified using distances (in basepairs) from regions on the linear genome to regions that define functional genomic annotations. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances have largely been ignored, yet have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.-->

@@

Furthermore... [??? You never describe the other data property - various levels of relationships]

Data level solutions mostly involve re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. [??? So, if the problem is solved, why do you bother with class imbalance at all?]

Evaluation level solutions include choosing suitable performance metrics for evaluating models. It has been shown that common metrics such as accuracy and AUROC (area under the receiver operating characteristic curve) are insensitive to class imbalance [@prati2011survey; @wei2013role]. A better measure to consider would be a composite metric involving all quadrants of the confusion matrix such as Matthews Correlation Coefficient (MCC). The suitability of MCC when evaluating models in imbalanced settings has been studied previously [@boughorbel2017optimal; @brodersen2010balanced]. Defined, the MCC is the correlation coefficient between the observed and predicted binary classifications. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. [??? So, if the problem is solved, why do you bother with it at all?]

[??? Carefully think on the "why" questions above. This paragraph answers it, connect it]
In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. The class imbalance issue was not addressed. Even worse, when performing cross-validation, [@mourad2016computational] used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance.

[??? Carefully think on the "why" questions above. This paragraph answers it, connect it]
A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases [?? Need to identify what you mean by "total number of cases." Is it just true positive divided by total number of possible boundaries ??]. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matrix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten] [?? MCC actually has all of the same components as AUC. AUC is a function of sensitivity and specificity which combined contain all quandrants ??]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<--
Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role]. 
-->

[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling [@chawla2002smote]. Under-sampling is performed without replacement from the majority class. On the other hand, over-sampling is performed by creating new synthetic observations using the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the feature vectors joining any/all of the k=5 (default) minority class nearest neighbors. The SMOTE algorithm can have many variations in the weightinh of over- and under-sampling. In this study we restricted the algorithm to 100% over-sampling and 200% under-sampling. That is, 1 new synthetic observation was to be created for each existing minority class data point, while 2 data points were simultaneously randomly selected from the majority class. This combination results in perfectly balanced classes. The *DMwR* R package was used to perform SMOTE (version 0.4.1).

@@

## Data construction for unbiased comparisons between resolution specific random forest classifiers
<!--## CELL LINE AND DATA [??? What data?] RESOLUTION SPECIFIC MODEL [??? What model?] CONSTRUCTION [??? Why?] !!!created new header-->

In total, we analyzed [??? Very boring] publically available Hi-C data for both the GM12878 and K562 cell lines, at four resolutions including 10 kb, 25 kb, 50 kb, and 100 kb. [??? But why? Problem is not defined] Given the varying levels of data resolution, we created a resolution-specific approach toward data constuction used for downstream modelling (Figure 1). This allowed us an unbiased way to compare performances of random forest classifiers between varying resolutions of Hi-C data. [??? Problem is not defined]

@@

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD boundaries, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). <!--This correlation [?? What correlation ??] with resolution was found to be significant, as evidenced by large intraclass correlation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.-->

**Table 1. Sparse TAD boundaries throughout the linear genome create high imbalance ratios irrespective of resolution.**

@@

<!--We then went on the evaluate the effect of class imbalance on model performance.--> In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity [?? Really confused by the constant use of granularity. Why is it needed ??] by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).
